---
title: "Understanding reactivity"
---

```{r}
#| label: "libs"
#| include: false
#| eval: true
#| echo: false

library(countdown)
```

## Introducing a simplified reactive app

The real magic of Shiny happens when you have an app with inputs and outputs 

. . .

and the two "react" to each other

```{r}
ui <- fluidPage(
  textInput(inputId = "name", 
            label = "What's your name?"),
  textOutput("greeting")
)

server <- function(input, output, session) {
  output$greeting <- renderText({
    paste0("Hello ", input$name, "!")
  })
}
```

## Looking for "name" and "greeting"

::: {.incremental}
- The ui `textInput()` is providing the information for the server `input$name` 
- And that is updating the object `output$greeting` in the server that updates the `textOutput("greeting")` in the ui
- The updates occur every time something is changed
:::

## The mental model

Unlike with R where updates like this occur when we run the code, here Shiny is being told *how it could* create the string if it needs to. The possibilities are:

::: {.incremental}
- This code is never run as the option is not used
- It might be run as soon as the app is launched
- It might be run once some time after the launch
- ... there are many possibilities
:::

. . .

:::{.callout-tip collapse=false appearance='default' icon=true}
## Tip!
Think of a Shiny app as providing recipes and not giving commands!
:::


## Introducing a new input

Four inputs were mentioned previously:

* `sliderInput()`  
* `selectInput()`  
* `textInput()`  
* `numericInput()`
. . .

And there are more! We are going to use `checkboxGroupInput()` in a new app.

## Building a new app

Starting again with a new script, typing `shinyapp` and then Shift + Tab add the code to `ui`, save and Run App

```{r title and checkbox}
 titlePanel("Basic Demo"), 
 checkboxGroupInput(
    inputId = "name1",
    label = "What are your favourite languages?",
    choices = c("SQL", "Python", "R")
  ),
```

## Adding an action button

Next we'll add an action button so, after making a selection, the user presses a button for the next thing to happen:

```{r}
  actionButton(inputId = "name2",
               label = "List the selections",
               icon = icon("question"))
```

. . .

:::{.callout-warning collapse=false appearance='default' icon=true}
## Don't forget the comma!
- In the ui the functions need a comma to separate
- The code will break unless it reads `checkboxGroupInput(...)`  **,**  `actionButton(...)`
:::

## Panels

Before we move to the interactive part of shiny we've currently got the layout where everything is appearing on the page as if it were a document.

## Panel functions

To give the app more structure that are more familiar we need to introduce three functions:

```{r}
#| code-line-numbers: 2|3|15
ui <- fluidPage(titlePanel("Basic Demo"), 
                sidebarLayout(
                sidebarPanel(
                    checkboxGroupInput(
                      inputId = "name1",
                      label = "What are your favourite languages?",
                      choices = c("SQL", "Python", "R")
                    ),
                    actionButton(
                      inputId = "name2",
                      label = "Selected",
                      icon = icon("question")
                    )
                ),
                mainPanel(textOutput(outputId = "name3"))
                )
)
```

## Tips

::: {.incremental}
- All three functions `sidebarLayout()`, `sidebarPanel()` and `mainPanel()` are needed to run the app
- If the copying and pasting loses the formatting, highlight and use `Ctrl+i` for indentation
- Alternatively install {styler} which can reformat selection or the entire script
:::

## `output$` and `input$` in the server

::: {.incremental}
- We now need to add in the server and have it "reactive" to what is selected in the ui
- The output will be simple in that we'll just show what is selected (a useful thing sometimes for debugging)
- Note in the `mainPanel()` the function `textOutput()` is used and that has a partnering function: `renderText()` in the server
:::

## server code

```{r}
server <- function(input, output, session) {
    observeEvent(input$name2, {
    output$name3 <- renderText(input$name1)
    })
}
```

Copy and paste this, save and run

## Exercise - renaming placeholders

Replace the names `name1`, `name2`, `name3` with more descriptive names, paying attention to where the repetitions occur

```{r}
#| eval: true
#| echo: false
countdown::countdown(minutes = 10,
                     color_border = "#005EB8",
                     color_text = "#005EB8",
                     color_running_text = "white",
                     color_running_background = "#005EB8",
                     color_finished_text = "#005EB8",
                     color_finished_background = "white",
                     margin = "0.9em",
                     font_size = "2em")
```

## `observeEvent()`

::: {.incremental}
- The app takes the input from the button in this example and until the button is pressed nothing will be printed.
- Once pressed any selection or de-selection is updated automatically.
- What about making the app require the button to be pressed before printing?
- For that we need a different function...
:::

## `bindEvent()`

`bindEvent()` can be used to execute only when a button is pressed. We'll replace the code 

```{r}
  observeEvent(input$name2, {
    output$name3 <- renderText(input$name1)
  })
```

with 

```{r}
output$name3 <- renderText({
  input$name1
}) |>
  bindEvent(input$name2)

```

:::{.callout-warning collapse=false appearance='default' icon=true}
## Remember your code will be different now
- In the last slide you will have changed `name1` and so on to something more readable!
- If names don't match you might not get an error just nothing prints.
:::

## The formal names - imperative and declarative 

Mastering Shiny explains this really well:

::: {.incremental}
- Imperative is where you issue a particular command and it's carried out immediately.
- This is what we are familiar with in analysis scripts.
- Declarative describes the goals and the software figures out how to achieved them without intervention.
:::

## Laziness

Although it's a strength of Shiny that declarative programming is lazy it does mean that code can have bugs but no errors.

If you recall when we changed `names1` to other names there was **no error** if these did not appear in both the ui and server.

This applies to functions so if don't pair them correctly:

```{r}
output$name3 <- renderPlot({ # should be renderText()
  input$name1
}) |>
  bindEvent(input$name2)

```

## Execution order

The familiar analysis scripts are run in the order of lines. 

Quarto reports for example, need to particularly have everything in the right order to run.

This isn't the case for Shiny! 

## [Exercise (from Mastering Shiny)](https://mastering-shiny.org/basic-reactivity.html#exercises-3)

Given this UI:

```{r}
ui <- fluidPage(
  textInput("name", "What's your name?"),
  textOutput("greeting")
)
```

Fix the simple errors found in each of the three server functions below. First try spotting the problem just by reading the code; then run the code to make sure you’ve fixed it.

```{r}
server1 <- function(input, output, server) {
  input$greeting <- renderText(paste0("Hello ", name))
}

server2 <- function(input, output, server) {
  greeting <- paste0("Hello ", input$name)
  output$greeting <- renderText(greeting)
}

server3 <- function(input, output, server) {
  output$greting <- paste0("Hello", input$name)
}
```

```{r}
#| eval: true
#| echo: false
countdown::countdown(minutes = 8,
                     color_border = "#005EB8",
                     color_text = "#005EB8",
                     color_running_text = "white",
                     color_running_background = "#005EB8",
                     color_finished_text = "#005EB8",
                     color_finished_background = "white",
                     margin = "0.9em",
                     font_size = "2em")
```

## Solution

```{r}

server1 <- function(input, output, session) {
  output$greeting <- renderText(paste0("Hello ", input$name))
}

server2 <- function(input, output, server) {
  greeting <- reactive(paste0("Hello ", input$name))
  output$greeting <- renderText(greeting())
}

server3 <- function(input, output, session) {
  output$greeting <- renderText({
    paste0("Hello ", input$name)
    })
}

```


## End session

### Acknowledgements

[Mastering Shiny](https://mastering-shiny.org/basic-app.html) by Hadley Wickham

[Building Web Apps with R Shiny](https://debruine.github.io/shinyintro/first-app.html) from [PsyTeachR](https://psyteachr.github.io/)
